% =============================================================================
% MAL Omics Manuscript
% COPDGene Multi-omics: Plasma Proteomics of Mechanically Affected Lung (MAL)
% =============================================================================
\documentclass[12pt, letterpaper]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath, amssymb}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{float}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{setspace}
\usepackage{xcolor}
\usepackage{soul}       % \hl{} for notes
\usepackage{lineno}
\usepackage{multirow}
\usepackage{rotating}

\doublespacing
\linenumbers

% Convenience for data-quality notes visible to reviewers
\newcommand{\datanote}[1]{\textcolor{red}{\textbf{[DATA NOTE: #1]}}}
\newcommand{\todo}[1]{\textcolor{blue}{\textbf{[TODO: #1]}}}

% =============================================================================
\begin{document}

\title{Plasma Proteomic Correlates of Mechanically Affected Lung and a
  Derived Risk Score for Emphysema Progression in COPDGene}

\author{
  \todo{Author list} \\
  \todo{Affiliations}
}
\date{\today}
\maketitle

% =============================================================================
\begin{abstract}
\textbf{Background.}
Mechanically Affected Lung (MAL) is a CT-derived continuous score
quantifying the volume of lung parenchyma under destructive mechanical
stress.  Although MAL predicts emphysema progression, the systemic
biological consequences detectable in peripheral blood are unknown.

\textbf{Methods.}
We profiled \(\sim\)4,979 plasma proteins using the SomaScan platform
in \(n = 1{,}306\) COPDGene participants.
Differential protein expression was assessed by multivariable linear
regression of each log\(_2\)-transformed protein against MAL score,
adjusting for age, sex, race, smoking, pack-years, CT scanner, and
ancestry principal components.  Gene-set enrichment analysis (GSEA)
and STRING network analysis characterised biological pathways.
An adaptive LASSO was used to construct a parsimonious Proteomic Risk
Score (PRS) for MAL.  The PRS was then tested for association with
emphysema progression (change in lung density, Phase~3 $-$ Phase~2)
using multivariable linear regression, and Akaike Information Criterion
(AIC) comparisons were performed for clinical, PRS-only, and combined
models.  Causal mediation analysis (medflex, natural effects framework)
evaluated whether the PRS mediates the pathway MAL $\to$ PRS $\to$
emphysema progression.

\textbf{Results.}
Forty-five proteins were differentially expressed with MAL at FDR $<$
0.05, led by Adiponectin (ADIPOQ, $\beta = 0.27$) and Troponin~T
(TNNT2, $\beta = 0.19$).  GSEA identified nine significant pathways,
predominantly reflecting metabolic suppression (fatty acid metabolism,
oxidative phosphorylation, glycolysis) and immune/inflammatory
activation (chemokine receptor signalling, antimicrobial peptides).
An adaptive LASSO selected 378 proteins that explained 15.7\% of MAL
variance in the test set.  The resulting PRS was not significantly
associated with emphysema progression ($\beta = 0.42$, 95\%~CI
[$-0.05$, $0.88$], $p = 0.079$) and did not improve model fit over
clinical variables alone ($\Delta$AIC $= +0.8$).  Causal mediation
showed that the PRS mediates $-0.6$\% of the MAL effect on progression
(Natural Indirect Effect $= -0.12$, $p = 0.97$), while the direct
effect of MAL was large and significant (Natural Direct Effect $= 19.7$,
$p = 0.004$).

\textbf{Conclusions.}
Plasma proteomics reflects the metabolic and inflammatory burden of
mechanically stressed lung.  However, the proteomic signature of MAL
does not sit causally between MAL and emphysema progression, nor does
it meaningfully enrich prediction beyond standard clinical variables,
suggesting that the mechanistic pathway from mechanical stress to
disease progression operates independently of the circulating
proteome.

\end{abstract}

\newpage

% =============================================================================
\section{Introduction}

Chronic obstructive pulmonary disease (COPD) is characterised by
progressive and irreversible emphysema driven in part by destructive
mechanical forces acting on lung parenchyma.  Mechanically Affected
Lung (MAL) is a continuous, CT-derived score (range 0.01--0.40) that
quantifies the fraction of lung volume under destructive mechanical
stress~\cite{todo_mal_ref}.  Higher MAL predicts emphysema progression
independently of baseline lung density and standard spirometric
measures, suggesting it captures a distinct pathophysiological process.

The systemic biological correlates of MAL that are detectable in
peripheral plasma are unknown.  Identifying plasma biomarkers of
mechanical lung stress would serve two complementary goals: (1)
understand the biological pathways by which mechanical forces damage the
lung, and (2) develop a blood-based test to enrich patient populations
for MAL and emphysema progression, enabling mechanistic studies and
clinical trials.

The COPDGene cohort~\cite{todo_copdgene_ref} has performed SomaScan
high-throughput plasma proteomics ($\sim$4,979 proteins) across more
than 5,000 participants.  We used this resource to: (i) map the
circulating protein correlates of MAL; (ii) characterise the implicated
biological pathways through gene-set enrichment and protein interaction
network analyses; (iii) build a parsimonious Proteomic Risk Score (PRS)
for MAL using adaptive LASSO penalised regression; and (iv) test whether
this PRS is associated with emphysema progression and whether it
mediates the causal effect of MAL on progression.

% =============================================================================
\section{Methods}

\subsection{Study Population}

Participants were drawn from the COPDGene study, a multi-centre
longitudinal cohort.  Plasma proteomic data (SomaScan), phenotype and
covariate data, and CT-derived MAL scores were linked by subject
identifier.  The analysis cohort comprised the complete cases with
non-missing values for all analysis variables (n = 1,306 after
three-way merge and complete-case filtering; see Table~\ref{tab:t1}).

\paragraph{Missing variables.}
\begin{itemize}
  \item \textbf{BMI}: No BMI column was identified in the phenotype
    file for this release (\texttt{grep("\^{}[Bb][Mm][Ii]")} returned
    no match).  BMI was therefore omitted from Table~1, the covariate
    set for differential expression, and all regression models.  When
    BMI data become available, it should be added as an unpenalised
    covariate.
  \item \textbf{FEV\textsubscript{1}\% predicted (FEV1pp)}: No column
    matching \texttt{FEV1pp\_P2} or related patterns
    (\texttt{FEV1pct}, \texttt{FEV1\_percent\_pred}) was found in the
    phenotype file.  FEV\textsubscript{1}\% predicted is therefore
    absent from Table~1.  The column \texttt{FEV1\_FVC\_post\_P2}
    (FEV\textsubscript{1}/FVC ratio) was present and is included.
\end{itemize}

\subsection{Proteomic Data}

SomaScan relative fluorescence unit (RFU) values for
$\sim$4,979 proteins were log\(_2\)-transformed prior to all analyses:
\[
  \tilde{x}_{ij} = \log_2\!\bigl(\max(x_{ij},\, 10^{-6})\bigr).
\]
Protein identifiers follow the pattern \texttt{log2\_X\{seqID\}} and
were mapped to HGNC gene symbols and UniProt targets via the SomaScan
annotation file (\texttt{metadat5k.csv}).

\subsection{Outcomes}

\begin{itemize}
  \item \textbf{MAL score} (\texttt{meanmal}): continuous, range
    0.01--0.40.
  \item \textbf{Emphysema progression}: change in CT lung density
    \(\Delta\rho = \rho_{\text{P3}} - \rho_{\text{P2}}\) (Hounsfield
    units, HU), stored as \texttt{Change\_lung\_density\_vnb\_P2\_P3}.
    Negative values indicate progression (density loss).
\end{itemize}

\subsection{Differential Protein Expression}

For each of the 4,979 proteins we fitted:
\[
  \tilde{x}_{\text{protein}} \sim \beta_0 + \beta_{\text{MAL}}\cdot
  \texttt{meanmal} + \boldsymbol{\gamma}^\top \mathbf{z} + \varepsilon,
\]
where \(\mathbf{z}\) comprises age, sex, race, pack-years, current
smoking status, CT scanner model, and ancestry principal components
(PC1--PC5).  The MAL coefficient \(\hat\beta_{\text{MAL}}\) and its
$t$-statistic were extracted for each protein.  Multiple testing was
controlled by the Benjamini--Hochberg false discovery rate (FDR);
proteins with FDR $< 0.05$ were considered differentially expressed.

\subsection{Gene-Set Enrichment Analysis}

All 4,979 proteins were ranked by their MAL $t$-statistic and submitted
to \texttt{fgsea}~\cite{todo_fgsea_ref} using gene sets from
\texttt{msigdbr}: Hallmark (H), KEGG Legacy (C2:CP:KEGG), and Reactome
(C2:CP:REACTOME) collections.  Minimum and maximum gene-set sizes were
15 and 500, respectively.  Pathways with Benjamini--Hochberg-adjusted
$p < 0.05$ were reported.

\subsection{STRING Network Analysis}

FDR-significant proteins were mapped to the STRING~v11.5 protein
interaction database (\texttt{STRINGdb} R package, human, score
threshold = 400) and the interaction network was plotted.

\subsection{Proteomic Risk Score for MAL}

\paragraph{Adaptive LASSO.}
We randomly split the cohort 70:30 into training (n = 914) and test
(n = 392) sets (\texttt{set.seed(2024)}).  A ridge regression pre-step
was applied to the residualised MAL outcome (after projecting out the
unpenalised covariates) to derive protein-specific adaptive weights:
\[
  w_j = \frac{1}{|\hat\beta^{\text{ridge}}_j|}.
\]
An L1-penalised regression (glmnet, $\alpha = 1$) with
\texttt{penalty.factor} $= w_j$ was then fitted using 10-fold
cross-validation; $\lambda_{\min}$ was selected.  Unpenalised
covariates (age, sex, race, smoking, pack-years, scanner, PC1--PC5)
were assigned \texttt{penalty.factor} $= 0$.

\paragraph{Post-selection OLS.}
Proteins with non-zero LASSO coefficients were refitted by ordinary
least squares (OLS) to obtain unbiased estimates, standard errors,
95\% confidence intervals, and $p$-values.

\paragraph{Proteomic Risk Score computation.}
For each subject $i$:
\[
  \text{PRS}_i = \sum_{j \in \mathcal{S}} \hat\beta_j^{\text{OLS}}
  \cdot \tilde{x}_{ij},
\]
where $\mathcal{S}$ is the selected protein set.  The PRS was
standardised to mean 0, SD 1 across the full cohort.

\subsection{Association of PRS with Emphysema Progression}

Three nested models were fitted by OLS:

\begin{enumerate}
  \item \textbf{Clinical model}: $\Delta\rho \sim$ age + sex + race +
    current smoking + pack-years + CT scanner
    \datanote{BMI omitted -- column absent from phenotype file. Add
    when available.}
  \item \textbf{PRS-only model}: $\Delta\rho \sim$ PRS +
    baseline lung density (\texttt{lung\_density\_vnb\_P2}) + CT scanner
  \item \textbf{Combined model}: $\Delta\rho \sim$ PRS + age + sex +
    race + current smoking + pack-years + CT scanner
    \datanote{BMI omitted.}
\end{enumerate}

Model fit was compared by AIC; a difference of $-2$ or more (lower) was
considered a meaningful improvement.  The PRS $\beta$ coefficient, 95\%
confidence interval, and $p$-value were reported from Model~3.

\subsection{Causal Mediation Analysis}

The causal pathway MAL $\to$ PRS $\to$ emphysema progression was
evaluated using the \texttt{medflex} R package (natural effects
framework)~\cite{todo_medflex_ref}.  An imputation model for the
mediator was specified as:
\[
  \Delta\rho \sim \texttt{meanmal} \times \text{PRS} + \text{covariates}
\]
using \texttt{neImpute()}, followed by \texttt{neModel()} with robust
(sandwich) standard errors.  The following quantities were estimated:
\begin{itemize}
  \item \textbf{Natural Direct Effect (NDE)}: effect of MAL on
    progression \emph{not} through the protein risk score
  \item \textbf{Natural Indirect Effect (NIE)}: effect of MAL on
    progression \emph{through} the protein risk score (i.e., mediation)
  \item \textbf{Total effect}: NDE + NIE
  \item \textbf{Proportion mediated}: NIE / Total
\end{itemize}
An analogous analysis was performed for each protein in the risk score
individually (Supplementary Table~S2).

\subsection{Software}

All analyses were performed in R (version $\geq$ 4.0) using:
\texttt{glmnet} (adaptive LASSO), \texttt{pROC} (AUC), \texttt{fgsea}
and \texttt{msigdbr} (GSEA), \texttt{STRINGdb} (network),
\texttt{medflex} (mediation), \texttt{gtsummary} (Table~1),
\texttt{ggplot2} and \texttt{ggrepel} (figures).  Full reproducible code
is available in the accompanying Jupyter notebook
\texttt{MAL\_omics/MAL\_omics\_proteomics.ipynb}.

% =============================================================================
\section{Results}

\subsection{Study Population}

After three-way linkage of proteomic, phenotype, and MAL data and
complete-case filtering, 1,306 participants were included.  Demographic
and clinical characteristics are summarised in Table~\ref{tab:t1}.

\paragraph{Note on missing clinical variables.}
FEV\textsubscript{1}\% predicted and BMI were not available in the
current phenotype file extract and are absent from Table~1.  Neither
variable was included as a covariate in any regression model.  Analysts
should verify whether these columns exist under alternative names in
other phenotype releases and update the analysis accordingly.

\subsection{Differential Protein Expression with MAL}

Of 4,979 proteins tested, \textbf{45 were significantly associated with
MAL} at FDR $< 0.05$ (Figure~\ref{fig:volcano}).  The strongest
positive associations (higher protein with higher MAL) were observed for
Adiponectin (\textit{ADIPOQ}, $\beta = 0.27$, FDR $= 4.9 \times
10^{-4}$) and Cardiac Troponin~T (\textit{TNNT2}, $\beta = 0.19$, FDR
$= 7.8 \times 10^{-4}$).  Additional proteins positively associated
with MAL included ADAMTS-5 (\textit{ADAMTS5}), NCAM-120
(\textit{NCAM1}), PAK7, MIA, and Chromogranin~B (\textit{CHGB}).

The strongest negative associations (lower protein with higher MAL)
were Prostasin (\textit{PRSS8}, $\beta = -0.12$), Aminoacylase-1
(\textit{ACY1}, $\beta = -0.26$), Nogo Receptor (\textit{RTN4R}, $\beta
= -0.15$), RET ($\beta = -0.18$), ADH4 ($\beta = -0.25$), and SFRP4
($\beta = -0.12$).  Full results for all 4,979 proteins are provided in
Supplementary Table~S1.

\subsection{Biological Pathway Enrichment (GSEA)}

Nine pathways reached significance (adjusted $p < 0.05$), summarised in
Figure~\ref{fig:gsea} and Table~\ref{tab:gsea_inline}:

\begin{table}[H]
\centering
\caption{Significant GSEA pathways (padj $< 0.05$). NES $< 0$ indicates
  depletion in high-MAL subjects.}
\label{tab:gsea_inline}
\small
\begin{tabular}{llrr}
\toprule
Collection & Pathway & NES & padj \\
\midrule
Hallmark  & Fatty Acid Metabolism             & $-1.76$ & 0.021 \\
Hallmark  & Oxidative Phosphorylation         & $-1.78$ & 0.021 \\
KEGG      & Glycolysis / Gluconeogenesis      & $-2.02$ & 0.018 \\
Reactome  & Biological Oxidations             & $-2.05$ & 0.007 \\
Reactome  & Metabolism of Amino Acids \& Derivatives & $-1.84$ & 0.013 \\
Reactome  & Diseases of Metabolism            & $-1.76$ & 0.032 \\
Reactome  & Phase I Functionalization of Compounds & $-1.97$ & 0.037 \\
Reactome  & Chemokine Receptors Bind Chemokines & $+2.02$ & 0.024 \\
Reactome  & Antimicrobial Peptides            & $+1.94$ & 0.032 \\
\bottomrule
\end{tabular}
\end{table}

Seven pathways were \textbf{depleted} in individuals with high MAL,
all reflecting core metabolic functions: energy metabolism (oxidative
phosphorylation, glycolysis), lipid catabolism (fatty acid metabolism),
amino acid handling, and xenobiotic processing (Phase~I, biological
oxidations).  Two pathways were \textbf{enriched}: chemokine receptor
signalling (NES $= 2.02$) and antimicrobial peptide activity (NES $=
1.94$), consistent with a shift toward an inflammatory state.

\subsection{STRING Protein Interaction Network}

FDR-significant proteins were mapped to the STRING~v11.5 human
protein--protein interaction network (score threshold $\geq 400$).  The
resulting network is shown in Figure~\ref{fig:string}.
\datanote{Describe network topology and key hubs once the figure is
reviewed: e.g., number of nodes, number of edges, main cluster
identities.}

\subsection{Proteomic Risk Score for MAL}

Adaptive LASSO selected \textbf{378 proteins} from 4,979 candidates.
On the held-out test set the score explained \textbf{15.7\%} of MAL
variance ($R^2 = 0.157$).  The top 20 proteins by absolute OLS
$\beta$-coefficient are listed in Table~\ref{tab:prs_top20}.  Strongest
positive contributors included ZPBP1 ($\hat\beta = 0.29$), GREM2
($\hat\beta = 0.19$), Complement~C7 ($\hat\beta = 0.17$), and
Collagen~Type~III (\textit{COL3A1}, $\hat\beta = 0.13$).  Strongest
negative contributors included Complement Factor~I
(\textit{CFI}, $\hat\beta = -0.23$), SEC13 ($\hat\beta = -0.15$), and
Spermine Synthase (\textit{SMS}, $\hat\beta = -0.13$).  Full weights
for all 378 proteins are provided in Supplementary Table~S2.

\subsection{PRS Association with Emphysema Progression}

Results of the three-model AIC comparison are shown in
Table~\ref{tab:aic}.  The clinical model achieved AIC $= 8{,}815$.
The PRS-only model was substantially \emph{worse} (AIC $= 8{,}848$,
$\Delta$AIC $= +33.1$), indicating that a proteomic-only model lacking
standard clinical predictors performs poorly.  Adding the PRS to the
clinical model provided essentially no improvement (AIC $= 8{,}816$,
$\Delta$AIC $= +0.8$ vs.\ clinical alone); this falls well short of
the $-2$ threshold for a meaningful gain.

In the combined model, the PRS coefficient was
$\hat\beta = 0.42$ HU per SD (95\%~CI [$-0.05$, $0.88$], $p = 0.079$),
indicating a modest, statistically non-significant trend.

The quartile analysis (Figure~\ref{fig:quartile}) shows the
PRS-associated effect size on emphysema progression across risk score
quartiles.

\subsection{Causal Mediation}

Results are reported in Table~\ref{tab:mediation}.  The Natural Direct
Effect of MAL on emphysema progression (the pathway \emph{not} through
the protein risk score) was large and statistically significant:
NDE $= 19.7$ HU ($p = 0.004$).  In contrast, the Natural Indirect
Effect (the mediated pathway through the PRS) was negligible:
NIE $= -0.12$ HU ($p = 0.97$).  The total effect was 19.6 HU, and the
proportion mediated was $-0.6$\%, effectively zero.

These results indicate that \textbf{the plasma proteomic signature of
MAL does not mediate the effect of MAL on emphysema progression}.  The
dominant pathway is direct and independent of the circulating proteome.

For individual proteins in the risk score, TREM2 showed the largest
proportion mediated (13.2\%, NIE $p = 0.017$), followed by SEZ6L
(16.8\%, $p = 0.025$) and RSPO1 (13.8\%, $p = 0.038$); however, all
individual indirect effects were small in absolute magnitude
(Supplementary Table~S3).

% =============================================================================
\section{Discussion}

We characterised the plasma proteomic landscape of mechanically stressed
lung in a large, well-phenotyped cohort.  Three principal findings
emerge.

\textbf{First}, MAL is associated with a distinct plasma protein
signature.  The dominant signals---elevated adiponectin and troponin~T,
suppressed metabolic enzymes (ADH4, ACY1)---are consistent with a
state of systemic metabolic stress and mechanical tissue damage.  The
positive association of ADAMTS5 (a matrix metalloproteinase involved in
aggrecan degradation) and the negative association of SFRP4 (a
Wnt-signalling inhibitor) suggest dysregulated extracellular matrix
remodelling.  Pathway analysis reinforces this picture: high-MAL
individuals show broad suppression of aerobic metabolism and
upregulation of innate immune pathways.

\textbf{Second}, despite selecting 378 proteins, the MAL PRS explained
only 15.7\% of MAL variance, and did not significantly predict
emphysema progression ($p = 0.079$) or improve AIC beyond clinical
variables alone ($\Delta$AIC $= +0.8$).  This suggests that the plasma
proteome reflects, but does not drive, the mechanical lung phenotype at
a level detectable above clinical noise.

\textbf{Third}, causal mediation analysis showed that the proteomic
signature mediates essentially none ($-0.6$\%) of the pathway from MAL
to emphysema progression.  The direct effect is overwhelmingly dominant
(NDE $= 19.7$, $p = 0.004$).  This implies that mechanical stress
causes progression through mechanisms that are not captured by the
peripheral proteome---perhaps local tissue processes (alveolar wall
stress, local inflammation) that do not produce detectable systemic
signals at this level of assay sensitivity.

\paragraph{Limitations.}
BMI and FEV\textsubscript{1}\% predicted were absent from the current
phenotype file extract and could not be included as covariates.  Their
absence may have introduced residual confounding in differential
expression and regression models.  Additionally, SomaScan measures
circulating aptamer-bound proteins; it does not capture local lung
tissue proteomics, which may be more directly relevant to the
mechanism.  The $R^2 = 0.157$ for MAL prediction indicates considerable
unexplained variance, likely reflecting the limitation of blood
proteomics as a proxy for a CT-derived structural measure.

% =============================================================================
\section*{Tables}

% --- Table 1 -----------------------------------------------------------------
\begin{table}[H]
\centering
\caption{Demographic and clinical characteristics of the study
  population ($n = 1{,}306$). Data are mean (SD) for continuous
  variables and $n$ (\%) for categorical variables. Emphysema
  progression is Phase~3 $-$ Phase~2 CT lung density (HU); negative
  values indicate progression.}
\label{tab:t1}
\small
\begin{tabular}{lc}
\toprule
Characteristic & Overall ($n = 1{,}306$) \\
\midrule
\multicolumn{2}{l}{\textit{Demographics}} \\
Age, years            & \todo{mean (SD)} \\
Sex, female           & \todo{n (\%)} \\
Race, White           & \todo{n (\%)} \\
\addlinespace
\multicolumn{2}{l}{\textit{Smoking}} \\
Current smoker        & \todo{n (\%)} \\
Pack-years            & \todo{mean (SD)} \\
\addlinespace
\multicolumn{2}{l}{\textit{Pulmonary function}} \\
FEV\textsubscript{1}\% predicted &
  \datanote{Column \texttt{FEV1pp\_P2} not found in phenotype file; omitted} \\
FEV\textsubscript{1}/FVC         & \todo{mean (SD)} \\
\addlinespace
\multicolumn{2}{l}{\textit{CT imaging}} \\
Lung density Phase~2 (HU)       & \todo{mean (SD)} \\
MAL score                        & \todo{mean (SD)} \\
Emphysema progression (HU)      & \todo{mean (SD)} \\
\addlinespace
\multicolumn{2}{l}{\textit{Anthropometric}} \\
BMI (kg/m\textsuperscript{2})   &
  \datanote{BMI column not found in phenotype file; omitted} \\
\bottomrule
\multicolumn{2}{l}{\footnotesize\textit{Note.} Values are populated
  from the \texttt{tbl\_summary()} output in the executed notebook.} \\
\multicolumn{2}{l}{\footnotesize\textit{Abbreviations:} MAL, Mechanically
  Affected Lung; FEV\textsubscript{1}, forced expiratory volume in 1~s;} \\
\multicolumn{2}{l}{\footnotesize FVC, forced vital capacity;
  HU, Hounsfield units.} \\
\end{tabular}
\end{table}

% --- Table 2: AIC ------------------------------------------------------------
\begin{table}[H]
\centering
\caption{AIC comparison of regression models for emphysema progression.
  $\Delta$AIC is relative to the clinical model. An improvement of
  $\leq -2$ is considered meaningful.}
\label{tab:aic}
\begin{tabular}{lrr}
\toprule
Model & AIC & $\Delta$AIC \\
\midrule
Clinical (age, sex, race, smoking, pack-years, scanner)\textsuperscript{a}
  & 8{,}815 & 0 \\
PRS only (PRS + baseline density + scanner)
  & 8{,}848 & $+33.1$ \\
Clinical + PRS
  & 8{,}816 & $+0.8$ \\
\bottomrule
\multicolumn{3}{l}{\footnotesize\textsuperscript{a}BMI omitted (column
  not available in current phenotype extract).} \\
\multicolumn{3}{l}{\footnotesize PRS $\hat\beta = 0.42$ HU/SD, 95\%~CI
  [$-0.05$, $0.88$], $p = 0.079$ in combined model.} \\
\end{tabular}
\end{table}

% --- Table 3: Mediation ------------------------------------------------------
\begin{table}[H]
\centering
\caption{Causal mediation analysis: MAL $\to$ Proteomic Risk Score
  $\to$ Emphysema Progression. Estimates from the natural effects model
  (medflex) with robust (sandwich) standard errors.}
\label{tab:mediation}
\begin{tabular}{lrrr}
\toprule
Parameter & Estimate (HU) & $z$ & $p$-value \\
\midrule
Natural Direct Effect (NDE)     & 19.72 & 2.92 & 0.004 \\
Natural Indirect Effect (NIE)   & $-0.12$ & $-0.04$ & 0.971 \\
Total effect                    & 19.60 & --- & --- \\
Proportion mediated             & \multicolumn{3}{c}{$-0.6$\%} \\
\bottomrule
\multicolumn{4}{l}{\footnotesize Covariates: age, sex, race, pack-years,
  current smoking, CT scanner, baseline lung density.} \\
\multicolumn{4}{l}{\footnotesize BMI omitted (not available). Results
  should be revisited when BMI data are added.} \\
\end{tabular}
\end{table}

% --- Table: Top 20 PRS proteins ----------------------------------------------
\begin{table}[H]
\centering
\caption{Top 20 proteins in the MAL Proteomic Risk Score by absolute
  OLS $\hat\beta$. CI, 95\% confidence interval.}
\label{tab:prs_top20}
\small
\begin{tabular}{llrrrr}
\toprule
Rank & Gene & Target & $\hat\beta$ & 95\%~CI & $p$ \\
\midrule
1  & ZPBP    & ZPBP1            & $+0.286$ & [$0.160$,\, $0.412$] & $<0.001$ \\
2  & CFI     & Factor I         & $-0.226$ & [$-0.398$,\, $-0.054$] & $0.010$ \\
3  & GREM2   & GREM2            & $+0.193$ & [$0.113$,\, $0.273$] & $<0.001$ \\
4  & C7      & Complement C7    & $+0.169$ & [$0.030$,\, $0.309$] & $0.017$ \\
5  & SEC13   & SEC13            & $-0.153$ & [$-0.254$,\, $-0.053$] & $0.003$ \\
6  & SERPINF2& $\alpha_2$-Antiplasmin & $+0.149$ & [$0.011$,\, $0.287$] & $0.035$ \\
7  & FSTL1   & FSTL1            & $-0.147$ & [$-0.309$,\, $0.015$] & $0.075$ \\
8  & APOH    & $\beta_2$-Glycoprotein~I & $+0.143$ & [$-0.049$,\, $0.335$] & $0.145$ \\
9  & COL3A1  & Collagen Type III & $+0.132$ & [$0.061$,\, $0.204$] & $<0.001$ \\
10 & C3      & C3a              & $+0.131$ & [$0.021$,\, $0.242$] & $0.020$ \\
11 & SMS     & Spermine Synthase & $-0.125$ & [$-0.179$,\, $-0.072$] & $<0.001$ \\
12 & MATN2   & MATN2            & $-0.125$ & [$-0.242$,\, $-0.008$] & $0.036$ \\
13 & GALNT11 & GLT11            & $-0.122$ & [$-0.225$,\, $-0.019$] & $0.021$ \\
14 & ROBO2   & ROBO2            & $-0.119$ & [$-0.220$,\, $-0.017$] & $0.022$ \\
15 & SEZ6L   & SEZ6L            & $+0.117$ & [$0.020$,\, $0.214$] & $0.019$ \\
16 & FBLN1   & Fibulin-1        & $-0.113$ & [$-0.278$,\, $0.051$] & $0.175$ \\
17 & ADAMTS5 & ADAMTS-5         & $+0.113$ & [$0.062$,\, $0.163$] & $<0.001$ \\
18 & CCL23   & MPIF-1           & $-0.112$ & [$-0.160$,\, $-0.063$] & $<0.001$ \\
19 & CREB3L4 & CR3L4            & $+0.108$ & [$0.047$,\, $0.169$] & $<0.001$ \\
20 & CNTN1   & Contactin-1      & $-0.107$ & [$-0.229$,\, $0.015$] & $0.086$ \\
\bottomrule
\end{tabular}
\end{table}

% =============================================================================
\section*{Figure Legends}

\begin{description}

\item[Figure 1 --- Study Design Schematic.]
  Overview of the analysis workflow. COPDGene participants with
  SomaScan plasma proteomics, CT-derived MAL scores, and longitudinal
  lung density measurements were studied. Differential expression
  identified proteins associated with MAL; pathway and network analyses
  characterised their biology; adaptive LASSO constructed a Proteomic
  Risk Score (PRS); and the PRS was tested for association with and
  mediation of emphysema progression.
  \datanote{Figure~1 is a conceptual schematic; to be drawn as a
  diagram (e.g., BioRender or vector graphics).}

\item[Figure 2 --- Protein Expression Landscape of MAL.]
  \textbf{(A) Volcano plot.} Each point is one protein ($n = 4{,}979$).
  The $x$-axis shows the MAL $\hat\beta$ coefficient; the $y$-axis shows
  $-\log_{10}(\text{FDR})$. Red: FDR $< 0.05$ and $\hat\beta > 0$
  (enriched with MAL); blue: FDR $< 0.05$ and $\hat\beta < 0$ (depleted
  with MAL); grey: not significant. Top 20 proteins by FDR are labelled.
  \textbf{(B) GSEA barplot.} Horizontal bars show the normalised
  enrichment score (NES) for all 9 significant pathways (padj $< 0.05$),
  coloured by direction and faceted by gene-set collection.
  \textbf{(C) STRING network.} Protein--protein interaction network
  for the 45 FDR-significant proteins (STRING~v11.5, score $\geq 400$).
  Node colour reflects the sign of the MAL association.

\item[Figure 3 --- Quartile Plot of Proteomic Risk Score.]
  Effect size (OLS $\hat\beta$ for the PRS predictor, HU/SD) with 95\%
  confidence intervals from separate regression models of emphysema
  progression on the PRS within each risk-score quartile. The $x$-axis
  shows PRS quartile (1 = lowest, 4 = highest); the $y$-axis shows the
  $\hat\beta$ coefficient. The dashed horizontal line marks the null.

\end{description}

% =============================================================================
\section*{Supplement}

\subsection*{Supplementary Tables}

\begin{description}
  \item[Table S1.] Complete differential expression results for all
    4,979 proteins (gene, target, $\hat\beta$, SE, $t$-statistic,
    $p$-value, FDR).  File: \texttt{MAL\_omics/mal\_de\_results.csv}.

  \item[Table S2.] Proteomic Risk Score weights for all 378 selected
    proteins ($\hat\beta_{\text{OLS}}$, 95\%~CI, $p$-value).  File:
    \texttt{MAL\_omics/mal\_risk\_score\_weights.csv}.

  \item[Table S3.] Per-protein mediation results for the 378 PRS
    proteins (NDE, NIE, total effect, proportion mediated, NIE
    $p$-value).  File: \texttt{MAL\_omics/mal\_mediation\_supp.csv}.
\end{description}

\subsection*{Supplementary Figures}

\begin{description}
  \item[Supplementary Figure A.] Forest plot of the top 30 FDR-significant
    proteins and their $\hat\beta$ coefficients for MAL association (HU
    per log\(_2\) RFU unit), with 95\% confidence intervals. Proteins
    are ordered by effect size.

  \item[Supplementary Figure B.] Forest plot of the same 30 FDR-significant
    proteins and their $\hat\beta$ coefficients for emphysema progression
    association (HU per log\(_2\) RFU unit, adjusted for baseline lung
    density and covariates), with 95\% confidence intervals.
\end{description}

% =============================================================================
\section*{Assumptions and Data Quality Notes}
\label{sec:data_notes}

The following issues were encountered during analysis and should be
addressed before finalising the manuscript:

\begin{enumerate}
  \item \textbf{BMI not found.} No column matching BMI was detected in
    \texttt{pheno\_MAL.csv}. The COPDGene data dictionary should be
    consulted for the correct column name (e.g.,
    \texttt{BMI\_P2}, \texttt{bmi\_P2}).  Once found, add
    \texttt{penalty.factor = 0} for BMI in the adaptive LASSO and
    include it in the clinical model and Table~1.

  \item \textbf{FEV\textsubscript{1}\% predicted not found.}  The
    expected column \texttt{FEV1pp\_P2} does not exist.  Possible
    alternatives: \texttt{FEV1pct\_pred\_P2}, \texttt{FEV1\_PP\_P2}.
    FEV\textsubscript{1}/FVC (\texttt{FEV1\_FVC\_post\_P2}) was found
    and used where spirometry was required.

  \item \textbf{Progressor definition.}  Emphysema progression is
    defined continuously as $\Delta\rho$ (HU). A binary definition
    (e.g., $\Delta\rho < -1.5$ HU) could be used for sensitivity
    analyses and would enable AUC/ROC-based model comparison.

  \item \textbf{R\textsuperscript{2} for MAL PRS.}  $R^2 = 0.157$ on
    the test set is modest.  This may reflect the inherent limitations
    of blood proteomics for a CT-derived structural phenotype, or
    insufficient power in the current split.  A bootstrap-stabilised
    estimate should be reported alongside the single test-set value.

  \item \textbf{Figure~1 schematic.}  A study design figure has not
    been automatically generated; it requires manual preparation (e.g.,
    BioRender).
\end{enumerate}

% =============================================================================
\bibliographystyle{unsrtnat}
\bibliography{references}
% Placeholder citations:
% \bibitem{todo_mal_ref}      Estépar et al. (MAL original paper)
% \bibitem{todo_copdgene_ref} Regan et al. COPDGene COPD 2010
% \bibitem{todo_fgsea_ref}    Korotkevich et al. bioRxiv 2021
% \bibitem{todo_medflex_ref}  Vansteelandt & Keiding, Epidemiology 2011

\end{document}
